Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-02-07 supplement.tex
update fig3 to use heterozygosity

@@

\subsection{Mutation versus recombination rates}

Prior studies primarily relied on the mutation clock in order to date events
such as population divergence, and uncertainty about the mutation rate
translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
Recent whole genome sequencing studies from non-pathological pedigrees have
typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
approximates the rate predicted by a mutational model from Gnomad
\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
demographic inference in this study. However, \moments differs from other
approaches by relying on a recombination clock, which has been argued to be
more robust to estimation than a mutational clock \citep{Moorjani2016-ur}. To
fit patterns of diversity which were inferred using the recombination map here,
mutation rates on the order of $2\times10^{-8}$ would be needed.
\aprcomment{This discrepancy between the mutation clock and recombination
clock\dots}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{
        \textbf{Predicted pairwise differentiation between contemporaneous populations.}
        $H_{i,j}$ is the predicted pairwise differentiaion between two populations,
        which we normalize by the observed pairwise diversity in the Mende sample.
        The single-origin model (A) shows larger deviations between observed and
        expected differentiation than the models with early human structure
        (B: continuousmigration, C: merger without stem migration, D: merger with 
        stem migration). In the merger models (C and D), while $F_{ST}$ is large
        between early stems due to the sharp bottleneck in one of those branches,
        the pairwise differences are comparable to the continuous-migration model.
        Stem 1E and Stem 1S have low differentiation, as the bottleneck in Stem 1
        (which they both split from) sharply reduces diversity by the time of the
        split. This has the effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}

>>>>

\subsection{What does this mean for expectation of morphology btw branches?}

\aprcomment{Tim to fill in, and edit section header}

@@

\subsection{Mutation versus recombination rates}
\label{sec:supp-mutation}

Prior studies primarily relied on the mutation clock in order to date events
such as population divergence, and uncertainty about the mutation rate
translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
Recent whole genome sequencing studies from non-pathological pedigrees have
typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
approximates the rate predicted by a mutational model from Gnomad
\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
demographic inference in this study. \moments differs from other approaches by
relying on a recombination clock, which has been argued to be more robust to
estimation than a mutational clock \citep{Moorjani2016-ur}.

From our recombination-rate-calibrated demographic models, we can estimate the
mutation rates needed to fit patterns of pairwise diversity. With our best-fit
model (merger-with-stem-migration,
Table~\ref{tab:supp-merger-with-stem-migration}), a mutation rate of
$1.33\times10^{-8}$ recovers the observed diversity within and between
populations, closely matching the average mutation rate estimated from the
Gnomad mutation model. The other inferred demographic models
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-without-stem-migration})
would require mutation rates of $1.75-1.90\times10^{-8}$ to match pairwise
diversity. \aprcomment{Because the merger-with-stem-migration model provides
the best fit to the LD statistics, the cSFS, \emph{and} the Gnomad mutation
rate, \dots}

\aprcomment{The differences in recovered mutation rates between models could
cut both ways: on one hand, it could help discriminate between models. One the
other hand, it should mutation rate estimation in this manner is sensitive to
the model parameterization, and we might expect it to be poorly constrained.
Did we just get lucky that our best-fit model also matches the Gnomad mutation
rate estimate for these same intergenic regions?}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
    \caption{
        \textbf{Predicted pairwise differentiation between contemporaneous
        populations.} $H_{i,j}$ is the predicted pairwise differentiaion
        between two populations per base-pair, using a mutation rate estimated
        from matching predicted to observed pairwise diversity from the
        intergenic data used in the fits (see Sec.~\ref{sec:supp-mutation}).
        The single-origin model (A) shows larger deviations between observed
        and expected differentiation than the models with early human structure
        (B: continuous migration, C: merger without stem migration, D: merger
        with stem migration). In the merger models (C and D), while $F_{ST}$ is
        large between early stems due to the sharp bottleneck in one of those
        branches, the pairwise differences are comparable to the
        continuous-migration model. Stem 1E and Stem 1S have low
        differentiation, as the bottleneck in Stem 1 (which they both split
        from) sharply reduces diversity by the time of the split. This has the
        effect of large $F_{ST}$ but small $H_{i,j}$.
    }
    \label{fig:supp-h12-panels}
\end{figure}
